---
title: "On Bicycles Stolen on Different Premises in Toronto from 2017 to 2023"
author: "Emily Su"
date: "16 January 2024"
date-format: "D MMMM YYYY"
abstract: "Toronto's population is increasing alongside the number of bicycle thefts across the city. We used data from Open Data Toronto to create a graph to discover any trends and differences between the number of bicycle thefts before 2020 and 2020 onwards. Our analysis found that there was a steep decline in the number of bicycle thefts after 2020 with a majority of thefts occurring in the premise type, Outside, in most years except for 2019, 2020, and 2021. Factors like the pandemic could have influenced the decline however, further investigation is needed on the demographic of people reporting."
thanks: "Data and code are available at: https://github.com/moonsdust/toronto-bicycles-stolen"
format:
  pdf:
    toc: true
number-sections: true
bibliography: references.bib
---
# Introduction
As Toronto's population increases, roads increasingly become packed with roads and bicycles have become one of the more favourable options for transportation, especially during the summer. However, this means that bicycle theft happens around Toronto and on different premises. With the pandemic starting in 2020, we are interested in seeing how the number of bicycle thefts has changed before 2020 versus 2020 onwards. 

In this paper, we will look at the number of bicycles stolen from 2017 to 2023 on different types of premises (Commercial (ex: Gas stations, construction sites, etc.), Outside (ex: Parking lots, roads, etc.), House, Apartment, Transit (ex: subway station), Educational, and Other) in Toronto. We then discuss our results and any implications of it.

# Process & Data {#sec-process}
In this section, we will discuss the process and some of the code used to analyze and clean the data from `Open Data Toronto` on Bicycle Thefts and create a graph from it [@citeOpenDataToronto]. See the footnote on the first page for the GitHub repository, which contains the data and all code pertaining to this paper. 

## Plan
The sketch showing how we expect the dataset and final graph to look can be found under `inputs/sketches/sketch.png` in the GitHub repository in the footnote of the first page. 

## Simulate
We will first add the preamble documentation and setup our workspace. We will use `R (programming language)` [@citeR], `tidyverse` [@citetidyverse], `janitor` [@citejanitor], `ggplot2` [@citeggplot2], and `knitr` [@citeknitr].
```{r}
#### Preamble ####
# Purpose: To create a graph of the number of bicycles stolen
# in Toronto on different premises from 2017 to 2023 by reading
# in data from Open Data Toronto.
# Author: Emily Su
# Email: em.su@mail.utoronto.ca
# Date: 16 January 2024
# Prerequisites: Know where the data is for bicycle thefts in Toronto. 
```

```{r}
#| echo: false
#| message: false

#| warning: false

#### Workspace setup #### 
## (Installing packages (only needs to be done once per computer))
# the four install.packages calls if this is the first time
# you opened this document

# install.packages("tidyverse")
# install.packages("janitor")
# install.packages("ggplot2")
# install.packages("knitr")


library(tidyverse) # Contains data-related packages
library(janitor) # Cleans datasets
library(ggplot2) # To make graphs
library(knitr) # To make tables
```

Now, we will create simulated data for the number of bicycle thefts on different Toronto premises from 2017 to 2023. These premises are the following: Commercial, Outside, House, Apartment, Transit, Educational, and Other.
```{r}
#| echo: false
#| message: false
#### Simulate data ####
set.seed(646)

number_of_years <- 7

simulated_data <-
  tibble(
    # We will repeat each premise name for <number_of_years> times 
    premise =
      c(rep("Commercial", number_of_years), 
        rep("Outside", number_of_years), 
        rep("House", number_of_years), 
        rep("Apartment", number_of_years),
        rep("Transit", number_of_years),
        rep("Educational", number_of_years),
        rep("Other", number_of_years)),
    # We will create a vector for the years 2017 to 2023
    # and repeat this <number_of_years> times for each premise 
    year = rep(c(1:number_of_years + 2016), number_of_years),
    # We will create <number_of_years> * 7 random values 
    # from a uniform distribution to get
    # the number of bikes of stolen for each year for each premise 
    num_bikes_stolen =
      round(runif(n = number_of_years * 7, min = 0, max = 49))
  )
```

```{r}
#| echo: false
#| message: false
#| tbl-cap: Simulated Data of the Number of Bicycle Stolen on Different Premises in Toronto (2017-2023)
# Creates table
head(simulated_data) |>
  kable()
```

## Acquire
In this step, we will first read in data from `Open Data Toronto` on Bicycle Thefts and then clean the dataset [@citeOpenDataToronto]. 
```{r}
#| echo: false
#| message: false
#### Read in the raw data ####
raw_bike_data <- read_csv(file = "../../inputs/data/bicycle-thefts.csv",
    show_col_types = FALSE)
```
Next, we will clean the column names of the raw data's dataframe, filter for the year 2017 to 2023, and select the following columns: `occ_year` and `premises_type`. We will then rename the column `occ_year` to `year` and create a new column called `num_bikes_stolen` to contain the number of bikes stolen for each `premises_type` for each `year`.
```{r}
#| echo: false
#| message: false
#### Clean ####
# Clean the column names
cleaned_bike_data <- clean_names(raw_bike_data)

# Select our desired columns 
cleaned_bike_data <-
  cleaned_bike_data |>
  filter(occ_year >= 2017 & occ_year <= 2023) |>
  select(
    occ_year,
    premises_type
  )

# Renaming occ_year column to year
cleaned_bike_data <-
  cleaned_bike_data |>
  rename( # Left side is the new name and the right side is the old name
    year = occ_year
  )

# Add new column, num_bikes_stolen and replace year column
# Referenced tutorial code from STA238
cleaned_bike_data <-
  cleaned_bike_data |>
  group_by(year, premises_type) |>
  mutate(num = n()) |> # count number of rows based on premise and year
  rename(num_bikes_stolen = num) |>
  unique() # This will filter out repeated rows
```
Now, we will run tests on the clean data to see if it passes the following conditions:

1. "year" does not contain years before 2017 and after 2023  

2. We have 7 types of "premises_type" 

3. num_bikes_stolen is greater than or equal to 0.
```{r}
# Tests 
# 1. "year" does not contain years before 2017 and after 2023
cleaned_bike_data$year |> min() == 2017
cleaned_bike_data$year |> max() == 2023

# 2. We have 7 types of "premise" 
cleaned_bike_data$premises_type |>
  unique() |>
  length() == 7

# 3. num_bikes_stolen is greater than or equal to 0.
cleaned_bike_data$num_bikes_stolen |> min() >= 0
```

```{r}
#| echo: false
#| message: false
# Export the cleaned data to a CSV file 
write_csv(
  x = cleaned_bike_data,
  file = "../data/cleaned_bike_data.csv"
)
```

## Explore
We will now create a table and graph for the cleaned data. 
```{r}
#| echo: false
#| message: false

# Import CSV contained the cleaned data
cleaned_bike_data <-
  read_csv(
    file = "../data/cleaned_bike_data.csv",
    show_col_types = FALSE
  )
```

```{r}
#| echo: false
#| message: false
#| tbl-cap: Number of Bicycles Stolen in Toronto on Different Premises from 2017 to 2023 
# Creates table
cleaned_bike_data |>
  kable()
```
```{r}
#| echo: false
#| message: false
#| label: fig-bikegraph
#| fig-cap: Number of Bicycles Stolen on Different Toronto Premises from 2017 to 2023
# Creates graph
cleaned_bike_data |>
  ggplot(mapping = aes(x = year, y = num_bikes_stolen, fill = premises_type)) +
  geom_bar(stat="identity") +
  ggtitle("Number of Bicycles Stolen on Different Premises in Toronto from 2017 to 2023") +
  theme_minimal() +
  labs(x = "Year", y = "Number of Bicycles Stolen", fill = "Type of Premise") +
  theme(legend.position = "bottom", plot.title = element_text(size = 9, face = "bold")) +
  scale_fill_viridis_d()
  # Referenced tutorial notes from summer STA238 2023 to obtain the code portion of geom_bar(stat="identity") and plot.title
```

# Results & Discussion {#sec-results}

We obtained data from Open Data Toronto on Bicycle Thefts that started in 2013 [@citeOpenDataToronto]. We then used the statistical programming language `R` [@citeR], `tidyverse` [@citetidyverse], and `janitor` [@citejanitor] to clean, tidy, and analyze the dataset. We followed the process (Plan, Simulate, Acquire, and Explore) to obtain a graph on the number of bicycles stolen during each year in Toronto for each type of premise (@fig-bikegraph) using `ggplot2` [@citeggplot2], and `knitr` [@citeknitr].

@fig-bikegraph from our results shows that 2018 had the most number of bicycle thefts with the most occurring Outside such as a parking lot and road. In contrast, the least number of bicycle thefts occurred in 2023 with the most thefts occurring at the premise type, outside. The number of bicycle thefts declined after 2020 when compared to the years before 2020. We can also see that in 2020 and 2021, the most number of bicycle thefts occurred at an apartment. We can predict that this could have been due to factors like the lockdowns and people staying at home more. This can also been seen in 2019 as well. For all years except for 2019, 2020, and 2021, the most bicycle thefts occurred outside.

However, this data only contained bicycle thefts that were reported and the result we would have gotten if all bicycle thefts were reported could look completely different from our current results. Some might have not reported due to various reasons such as historical discrimination from the police, the police station being located too far, or not having access to a phone. Further investigation is needed on the demographics (such as income) of people reporting bicycle theft.

\newpage

# References {#sec-references}